---
title: "Q-sort methodology proof of concept for Tom Lang"
output:
  html_document:
    df_print: paged
---

by Martin Brown, Martin.Brown@state.or.us, +1 503-229-5502

## Introduction

This R markdown notebook is a demonstration that a ["qsort" analysis](https://en.wikipedia.org/wiki/Q_methodology) is technically feasible here at Oregon DEQ.  DEQ's Tom Lang is interested in using q sorts to understand attitudes towards waste management and materials management in western Oregon counties.

I will try to:

* import some qsort field data which Tom Lang has entered via [PQMethod](http://schmolck.org/qmethod/) software
* utilize that data in a pre-programmed q sort, performed by the R package [qmethod](https://journal.r-project.org/archive/2014-2/zabala.pdf)
* display standard output (plots and data tables) produced by the q sort method.
* alter some of the standard output for better comprehension (e.g. by adding full statement labels)

In this session I will not attempt to understand or interpret the content of the q sorts.

## Setting up workspace

Loading required R packages...

```{r}
library(tidyverse)
library(qmethod)
library(knitr)
library(rmarkdown)
library(kableExtra)
library(sjPlot)
#library(plotly)
```

## Importing sample data

```{r}
# here is the score data itself
my_better_scores_1 <-
  import.pqmethod(
    file ="source_data/Tillamoo.dat"
  ) 
short_statement_names_1 <- row.names(my_better_scores_1)
my_better_scores_1a <-
  data.frame(my_better_scores_1, sta=short_statement_names_1)
my_better_scores_2 <-
  import.pqmethod(
    file="source_data/Sort_2_T.dat"
  ) 
short_statement_names_2 <- row.names(my_better_scores_2)
my_better_scores_2a <-
  data.frame(my_better_scores_2, sta=short_statement_names_2)

my_better_scores <-
  bind_rows(
    my_better_scores_1a %>% mutate(timeFrame="immediate"), 
    my_better_scores_2a %>% mutate(timeFrame="future")
    ) %>%
  select(-X16A)

# here is the full text of each statement, as a 1-field
# data frame
my_better_statements_1 <-
  read.delim(
    file="source_data/Tillamoo.sta",
    header=FALSE,
    col.names = c("long_statement")
    )

my_better_statements_2 <-
  read.delim(
    file="source_data/Sort_2_T.sta",
    header=FALSE,
    col.names = c("long_statement")
    )

my_better_statements <- 
  bind_rows(my_better_statements_1, my_better_statements_2) %>%
  unique()

my_statement_key =
  bind_cols(
    select(my_better_scores_1a, sta),
    my_better_statements
  )

my_data <-
  bind_cols(
    my_better_scores,
    data.frame(
      shortSta=c(short_statement_names_1,short_statement_names_2)
    )
  )

my_data_long <-
  pivot_longer(
    my_data,
    names_to = "respondent",
    values_to = "score",
    X1A:X15A
  )
```

## Scores for each statement

```{r}
my_descriptives <-
  my_data_long %>%
  group_by(timeFrame, shortSta) %>%
  summarise(
    min=min(score),
    med=median(score),
    max=max(score),
    mean=mean(score),
    sd=sd(score)
  ) %>% ungroup()

my_descriptives_2 <-
  left_join(
    my_descriptives,
    my_statement_key,
    by = c("shortSta"="sta")
  )
```


## Running an analysis

For simplicity I will run this asking for only 2 factors.

```{r}
my_analysis <-
  qmethod(dataset=my_better_scores_1, nfactors=5, rotation = "cluster")
```

## Standard outputs

At this point the R object "my_analysis" is storing the results.  R can output this in various ways.

A summary statement...

```{r}
summary(my_analysis)
```

A standard plot..

```{r fig.height=10, fig.width=6.5}
plot(my_analysis)
```

A standard printout... (may be long)
```{r}
print(my_analysis, length=28)
```

## Some attempt at customizing output

Here I will draw out the z-scores for each factor, label them with the long statement names.
```{r, fig.height=10, fig.width=6.5}
my_own_plot_data <-
  my_analysis$zsc 
short_statement <- row.names(my_own_plot_data)
my_own_plot_data_2 <-
  bind_cols(
    data.frame(short_statement=short_statement), 
    my_own_plot_data
  )
my_own_plot_data_3 <-
  full_join(
    my_own_plot_data_2,
    my_statement_key,
    by="short_statement"
  )
my_own_plot_data_4 <-
  pivot_longer(
    data=my_own_plot_data_3,
    names_to = "factor",
    values_to = "z_score",
    zsc_f1:zsc_f2
  )
ggplot()+
  ggtitle("Alternate z score chart with custom labels")+
  geom_point(
    data=my_own_plot_data_4,
    aes(y=z_score, x=long_statement, color=factor, shape=factor)
  )+
  theme(
    axis.text.x=element_text(size=5, angle=90, hjust=1, vjust=0.5)
  )
```

Ok, I think I've shown we can produce output.  Now, to make it mean something is the next task.
